---
title: "About This Website"
toc: FALSE
format:
  html:
    code_folding: show
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  echo = TRUE
)
```

This is an [R Markdown website](http://rmarkdown.rstudio.com/rmarkdown_websites.html), created by [Alex Kaizer](https://www.alexkaizer.com) for a short course on adaptive and Bayesian clinical trial designs for the 2024 WNAR/IMS/Graybill Conference hosted at Colorado State University.

In the short course, we'll actively use several packages which are listed below. Code to install the packages follows, and version information about each of these packages and their dependencies is at the bottom of this page.

For step-by-step instructions on installing `brms` and Stan, the following learnB4SS provides nice steps on their [GitHub website](https://learnb4ss.github.io/learnB4SS/articles/install-brms.html). 

Installing packages from CRAN

```{r, eval=FALSE}
# main packages
install.packages(c("brms","bayestestR","rpact","carat","kableExtra","blindrecalc","gsDesign"))
```

Versions of key packages:

```{r}
devtools::session_info(c("brms","bayestestR","rpact","carat","kableExtra","blindrecalc","gsDesign"))
```

